Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-16 Ben_Manuscripts/hdphmm/Draft.tex
Comments.

  Appropriate time series modeling of complex diffusion in soft matter systems on the
  microsecond time scale can provide a path towards inferring transport mechanisms and
  predicting bulk properties characteristic of much longer time scales. In this work 
  we apply the sticky hierarchical Dirichlet process hidden Markov model (HDP-AR-HMM) 
  to solute center-of-mass trajectories generated from long molecular dynamics 
  simulations in a cross-linked inverted hexagonal phase lyotropic liquid crystal 
  (LLC) membrane in order to automatically detect a variety of solute dynamical states
  which can be further explained in terms of solute-membrane interactions.

  We group the states identified by the HDP-AR-HMM into clusters based on multiple metrics
  aimed at distinguishing solute behavior based on their fluctuations, dwell times
  in each state and position within the inhomogeneous membrane structure. We analyze
  prevalent clusters in order to relate their parameters to physical interactions between 
  solutes and the membrane.

  Along with parameters of individual states, the HDP-AR-HMM simultaneously infers a transition
  matrix which allows us to stochastically propagate solute behavior from all of the 
  independent trajectories onto much longer time scales while still preserving the 
  qualitative behavior characteristic of the MD trajectories. This affords a direct 
  connection to important macroscopic observables used to characterize performance like
  solute flux and selectivity.

  Overall, this work provides an effective way to simultaneously identify transport 
  mechanisms in nanoporous materials and project complex diffusive behavior on
  long time scales.  This enhancement to our understanding of the diverse range 
  of solute behavior allows us to hypothesize design changes 
  to LLC monomers aimed towards controlling the rates of solute passage, thus improving 
  the selective performance of LLC membranes.

@@

  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams. Seawater and
  bracksih water desalination, separation of organic micropollutants from 
  municipal water supplies, recovering valuable dissolved species from 
  waste stream and creating breathable barriers to harmful contaminants are just a
  subset of the broad range of applications for selective membrane separations.~\cite{fritzmann_state---art_2007,fonseca_couto_critical_2018,dischinger_evaluation_2019,ramaseshan_functionalized_2006}
  While researchers often focus on membrane permeability, many argue that
  it may be more beneficial to focus on membrane selectivity in order to 
  reduce the costs of commercial nanofiltration and reverse osmosis.~\cite{werber_materials_2016}

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables. If one could apply molecular dynamics 
  (MD) simulations to explain important experimental metrics such as solute flux and 
  selectivity in terms of detailed chemically dependent solute motion, then we can more
  easily bridge the disconnect between theory and application. In many systems, it's 
  possible to achieve this by calculating diffusion constants. 
  % BJC4: even if we could calculate diffusion constants, wouldn't there still be value 
  % in understanding chemical detail?
  % BJC4: It might not be a good idea to blame our need for this modeling on non-linear MSDs, 
  % especially since our MSDs appear to be at or nearing a linear regime.
  % BJC4: seems like we should be addressing the problem as short time scale subdiffusive 
  % dynamics that dictates measureable long time scale diffusion. Although curves become linear,
  % the slopes should be a product of the short time scale motion.
  Unfortunately, for 
  systems which exhibit complex short timescale subdiffusive dynamics, one cannot 
  reliably compute diffusion constants since solute mean square 
  displacements (MSDs) are non-linear on the timescales accessible to molecular simulation.
  Therefore, we are in need of a more flexible approach for modeling solute dynamics
  which makes less assumptions about the solute's short time scale behavior.

  In this work, we are particularly interested in the design of lyotropic liquid 
  crystal (LLC) monomers, a class of amphiphilic molecules whose ordered self-assembled
  phases can be cross-linked into mechanically strong membranes capable of highly 
  selective separations. Inverted hexagonal, H\textsubscript{II}, phase LLC membranes
  are characterized	by hexagonally packed, uniform-sized and straight pores, an ideal
  geometry for high throughput transport. Their pores are lined with the LLC monomer's
  functional groups which can potentially be designed to interact with solutes in a 
  chemically-specific manner. It would be extremely useful to experimentalists if we 
  could connect LLC monomer design with measurable quantities. This goal is complicated
  by the non-Brownian hopping and trapping behavior induced by the membrane's diverse
  topology and inhomogeneous structure.~\cite{coscia_understanding_2019,coscia_chemically_2019}

  We have already progressed our understanding of LLC monomer design in our previous work
  by building stochastic time series models based on the chemical intuition gained
  from qualitative MD studies of solute transport in an H\textsubscript{II} phase LLC
  membrane.~\cite{coscia_chemically_2019,coscia_capturing_2020} We identified three 
  classes of trapping behavior which result in short time scale subdiffusion. In our 
  first approach to stochastically model this behavior, we treated the system's dynamics
  as a sequence of anti-correlated hops between power law distributed periods of 
  entrapment, a framework called subordinated fractional Brownian or L\'evy
  motion.~\cite{thiel_weak_2014,teuerle_modeling_2013} In a second approach, we treated 
  solute motion as a Markov state model~\cite{pande_everything_2010} with state-dependent
  dynamics, where we parameterized the state transition probabilities between each of
  eight discrete states, defined by the observed trapping mechanisms, as well as the 
  solute dynamics within each of these states. We showed how one could use realizations
  of any stochastic model like these in order to predict macroscopic flux and selectivity.

  Although powerful and insightful, our previous work required considerable human effort
  in order to analyze the MD trajectories and to hypothesize stochastic models which
  matched the behavior of solutes. It would be significantly more useful and broadly 
  applicable if we could design an approach which automatically distinguishes and 
  parameterizes varied solute dynamics and leaves a facile way to generate an ensemble
  of stochastic trajectory realizations that can be used to predict macroscopic transport
  properties.

@@

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but independent observations from 
  the emission distributions associated with each hidden state. Transitions 
  between states are mathematically defined by a matrix of transition probabilities.
  Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm is the most commonly used way to determine the hidden
  state sequence and the underlying parameters of a standard HMM.~\cite{baum_maximization_1970}
  It also possible to achieve the same results using a Bayesian 
  framework.~\cite{scott_bayesian_2002,jasra_markov_2005} One can place priors
  on the transition probabilities and parameters of the hidden states' emission
  distributions in order to iteratively infer their values and to estimate the hidden state
  sequence. HMMs have been used in a broad range of applications including speech
  recognition, weather forecasting and bioinformatics.~\cite{juang_hidden_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}. In molecular simulations, HMMs have been
  used to characterize molecular kinetics in order to determine the equilibrium 
  distribution of various observables.~\cite{noe_probability_2008}

  By placing a hierarchical Dirichlet process (HDP) prior on the transition 
  probabilities, it is possible to extend HMMs to an unknown number of hidden
  states, thus allowing the model's complexity to be driven by the complexity
  of the data.~\cite{teh_hierarchical_2006} We will refer to this as the HDP-HMM,
  although in literature it is often called the infinite hidden Markov 
  model.~\cite{beal_infinite_2002} The HDP-HMM elegantly overcomes the main
  shortcoming of standard HMMs, the need to choose the number of states 
  \textit{a priori}, making its use more advantageous in a diversity of 
  applications including economic data and single particle trajectory 
  analysis.~\cite{shi_identifying_2016,hines_analyzing_2015}

  One issue with the original formulation of the HDP-HMM, is that it does 
  not distinguish self-transitions and transitions between different states.
  Practically, this can result in rapid switching between states and inferred
  state sequence representations that do a poor job of describing the 
  observations. Fox et al. addressed this issue by adding a `sticky' 
  parameter.~\cite{fox_sticky_2007} The sticky HDP-HMM is useful in situations
  where the user has prior knowledge that state persistence is likely.
  For example, in the case of speaker diarization, it is unlikely that a 
  conversation between two people would rapidly switch between speakers.~\cite{fox_sticky_2011}
  This is also important for our work where we expect state persistence due
  to solute entrapment.

@@

  To our knowledge, this is a novel application and analysis of the sticky 
  HDP-AR-HMM. Typical applications center around determining the number of
  states, the state sequence and occasional analysis of the emission 
  distributions. The addition of temporal correlation to the inference 
  procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015,hamada_modeling_2016}
  Not only do we use the HDP-AR-HMM to segment MD time series into distinct
  dynamical modes, but we cluster and analyze the temporally-correlated 
  parameters of the model in order to learn physical mechanisms and describe
  solute kinetics. Further, we use the model parameters to generate new 
  time series trajectories which enable projections of MD simulations on 
  much longer time scales.

@@

  We ran all MD simulations and energy minimizations using GROMACS 2018.
  ~\cite{bekker_gromacs:_1993,berendsen_gromacs:_1995,van_der_spoel_gromacs:_2005,hess_gromacs_2008}  
  Our python implementation of the HDP-AR-HMM algorithm is available online at \\
  \texttt{https://github.com/bencoscia/hdphmm}. All other post-simulation 
  trajectory analysis tools are available online at
  \texttt{https://github.com/shirtsgroup/LLC\_Membranes}.

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10 \% water by weight. Approximately one third of the water molecules 
  occupy the tail region with the rest near the pore center.

@@

  We let each solute undergo 5 $\mu$s of MD simulation. We used a time step of 2 fs
  at a pressure of 1 bar and temperature of 300K controlled by the Parrinello-Rahman 
  barostat~\cite{parrinello_polymorphic_1981} and velocity rescale 
  thermostat~\cite{bussi_canonical_2007} respectively. We recorded configurations every 0.5 ns.

@@

  We implemented the sticky HDP-AR-HMM in Python 
  (\texttt{https://github.com/bencoscia/hdphmm}) which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_nonparametric_2009} Since we do not incorporate
  any novel methodology to the HDP-AR-HMM inference procedure, we refer the 
  interested reader to an excellent and in-depth description of the formulation of
  the HDP-AR-HMM and related models.~\cite{fox_bayesian_2010}

  %\subsubsection*{Parameterizing the Hidden States}\label{method:var_params}
  \textit{Parameterizing the Hidden States}: We describe the dynamics of each 
  state visited by solutes in our MD simulations using a first order vector 
  autoregressive (VAR(1)) model. In general, a VAR($r$) process is characterized by a
  vector of observations in a time series that are linearly dependent on $r$ previous
  values of the time series vector:
  \begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  \label{eqn:var}
  \end{equation}
  Previous observations are weighted by coefficient matrices, $A_i$. The VAR($r$) 
  process is further characterized by a shift in the mean of each dimension by the
  vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
  We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  covariance, $\Sigma$. We limited our analysis to an autoregressive order of $r=1$.
  This means that we only parameterize $A_1$. To simplify notation, we will just
  call it $A$. We used conjugate matrix-normal inverse-Wishart (MNIW) prior on parameters 
  $A$ and $\Sigma$ and a Gaussian prior on $\mathbf{c}$ in order to infer their 
  values.~\cite{fox_nonparametric_2009}

@@

  We reduced the parameter space using agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage 
  criteria, which works to minimize the sum of the squared differences within all
  clusters.~\cite{ward_hierarchical_1963} We elected to choose the number of clusters
  rather than the distance threshold. For our data, non-parametric methods such as 
  Dirchlet process Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013}
  tend to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information), which could cause the clustered model to combine 
  distinct behavior into a non-physical hybrid state.

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the distance from the 
  pore center and the self-transition probabilities, $T_{ii}$, of each state. $A$ 
  and $\Sigma$ capture the correlation and size of fluctuations. $T_{ii}$ is directly
  related to the average time spent in each state. The distance from the closest pore 
  center is likely meaningful since the membrane is radially inhomogeneous from each 
  pore, transitioning from hydrophilic pores to hydrophobic tails. However, the 
  channels are isotropic along the pore axis so we do not in means in the $z$ dimension.
  We normalized each clustering dimension by its variance and shifted the mean to zero.

  To take advantage of the system's cylindrical symmetry, we converted the parameters that
  are directly related to solute position into cylindrical coordinates before clustering. 
  We replaced the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ with $\lambda_x$ + 
  $\lambda_y$ and $\lambda_x^2$ + $\lambda_y^2$ respectively, which are invariant with 
  rotation around the $z$ axis, in both cases reducing the dimensionality from 3 to 2.  
  We clustered on squared means in $r$, $\mu_r^2 = \mu_x^2 + \mu_y^2$, rather than
  $\mu_x$ and $\mu_y$ individually, with the square $\mu_r$ converting it into the 
  same units as $A$ and $\Sigma$.

@@

  %BJC4: Added
  There are alternative ways to cluster the data dependent on one's goals. Our goal is
  to derive the simplest set of parameter which adequately describe solute motion. 
  However, one may want a more detailed understand of specific dimensions of the 
  clustering dataset. In this case one can cluster on one or more dimensions independently
  and then enumerate the possible clusters. This would ensure very little overlap of
  clusters but may result in a large number of them. Once one has achieved a better 
  understanding of solute-membrane interactions, perhaps by using the HDP-AR-HMM, it may be
  beneficial to incorporate other physically motivated variables into the clustering
  procedure. For example, one can gain a thorough understanding of the dynamics of
  different kinds of trapping behavior related to hydrogen bonding and sodium ion association.

  %\subsubsection*{Choosing the Number of Clusters}

@@

  \subsection{Predicting Mean Squared Displacement}\label{method:realizations}
  %BJC4: reworked this section
  \textit{Generating Stochastic HDP-AR-HMM Trajectory Realizations}: We used two 
  different methods to stochastically generate trajectories:
  \begin{enumerate}[label={Method \theenumi :}, leftmargin=3.5\parindent]
    \item For each independent MD trajectory, we generated stochastic trajectory
    realizations using the finalized parameters of the associated HDP-AR-HMM. This 
    results in 24 separate ensembles of trajectories for each solute studied. 
    \item Using the clustered parameter sets, we generated a single ensemble
	of trajectories. Realizations are representative of the combined behavior
	of all solute trajectories.
  \end{enumerate}

@@

  For MD trajectories, we bootstrapped according to Method 2.

@@

  Solute flux, as written, can be expressed in terms of the product of solute 
  diffusivity, $D_i$, within the membrane and a sorption coefficient, $K_i$, 
  that quantifies the ability of the solute to move between the bulk solution
  and and the membrane:
  \begin{equation}
    J_i = \frac{D_i K_i}{l}\Delta C
  \end{equation}
  where $l$ is the membrane thickness. For our purposes, $K_i$ and $l$ are 
  irrelevant since we do not study pore entry effects or flux across finite
  length pores. Therefore, in order to calculate selectivity, with respect
  to quantities we can study with our MD simulations, it is only necessary that
  we compare the ratio of diffusion constants:
  \begin{equation}
    S_{ij} = \frac{D_i}{D_j}
    \label{eqn:selectivity_diffusivity}
  \end{equation}
  It can be shown on time scales where motion is uncorrelated, that the diffusion
  constant is linearly proportional to the slope of the MSD curve:~\cite{einstein_investigations_1956}
  \begin{equation}
    \overline{z^2(\tau)} = 2 n D t
    \label{eqn:msd_D}
  \end{equation}
  where $n$ is the number of spatial dimensions (1 in our case). Since the MSD
  of our models enters a linear regime, we can estimate diffusion constants 
  based on the slope of the linear region of the predicted MSD curves.

@@

  \textit{Hydrogen bond lifetimes}: We estimated the lifetime of hydrogen bonds by 
  recording the length of sequential frames where solutes remained hydrogen bonded.
  We still counted sequences where hydrogen bonds were broken for a single frame 
  before reforming. Consistent with our previous work, we reported the 95th 
  percentile of hydrogen bond lifetimes since their distribution is not Gaussian
  and to emphasize longer trapping periods.~\cite{coscia_chemically_2019}

>>>>

  Appropriate time series modeling of complex diffusion in soft matter systems on the
  microsecond time scale can provide a path towards inferring transport mechanisms and
  predicting bulk properties characteristic of much longer time scales. In this work 
  we apply the sticky hierarchical Dirichlet process hidden Markov model (HPD-AR-HMM) 
  %MRS: might be confusing to have the AR in there when the words ``Auto-regressive'' 
  %don't appear; then they would try to figure out where the AR is. 
  to solute center-of-mass trajectories generated from long molecular dynamics 
  simulations in a cross-linked inverted hexagonal phase lyotropic liquid crystal
  (LLC) membrane in order to automatically detect a variety of solute dynamical states
  which can be further explained in terms of solute-membrane interactions.

  We group the states identified by the HDP-AR-HMM into clusters based on multiple metrics
  aimed at distinguishing solute behavior based on their fluctuations, dwell times
  in each state and position within the inhomogeneous membrane structure. We analyze
  prevalent clusters in order to relate their parameters to physical interactions between 
  solutes and the membrane. 
  Along with parameters of individual states, the HDP-AR-HMM simultaneously infers a transition
  matrix which allows us to stochastically propagate solute behavior from all of the 
  independent trajectories onto much longer time scales while still preserving the 
  qualitative behavior characteristic of the MD trajectories. This affords a direct 
  connection to important macroscopic observables used to characterize performance like
  solute flux and selectivity.

  %MRS: we have to decide if ``effective'' is the right word. Alternative is ``promising''.
  Overall, this work provides an effective way to simultaneously identify transport 
  mechanisms in nanoporous materials and project complex diffusive behavior on
  long time scales.  
  %MRS: simplify by removing nominalizations.
  %This enhancement to our understanding 
  Our enhanced understanding
  of the diverse range 
  of solute behavior allows us to hypothesize design changes 
  to LLC monomers aimed towards controlling the rates of solute passage, thus improving 
  the selective performance of LLC membranes.

@@

  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams. Seawater and
  brackish water desalination, separation of organic micropollutants from 
  municipal water supplies, recovering valuable dissolved species from 
  waste stream and creating breathable barriers to harmful contaminants are just a
  subset of the broad range of applications for selective membrane separations.~\cite{fritzmann_state---art_2007,fonseca_couto_critical_2018,dischinger_evaluation_2019,ramaseshan_functionalized_2006}
  While researchers often focus on membrane permeability, many argue that
  it may be more beneficial to focus on membrane selectivity in order to 
  reduce the costs of commercial nanofiltration and reverse osmosis.~\cite{werber_materials_2016}

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables. If one could apply molecular dynamics 
  (MD) simulations to explain important experimental metrics such as solute flux and 
  selectivity in terms of detailed chemically dependent solute motion, then we can more
  easily bridge the disconnect between theory and application. In many systems, it is 
  possible to achieve this by calculating diffusion constants. 
  % BJC4: even if we could calculate diffusion constants, wouldn't there still be value 
  % in understanding chemical detail?
  % BJC4: It might not be a good idea to blame our need for this modeling on non-linear MSDs, 
  % especially since our MSDs appear to be at or nearing a linear regime.
  % BJC4: seems like we should be addressing the problem as short time scale subdiffusive 
  % dynamics that dictates measurable long time scale diffusion. Although curves become linear,
  % the slopes should be a product of the short time scale motion.
  Unfortunately, for 
  systems which exhibit complex short timescale subdiffusive dynamics, one cannot 
  reliably compute diffusion constants since solute mean square 
  displacements (MSDs) are non-linear on the timescales accessible to molecular simulation.
  Therefore, we are in need of a more flexible approach for modeling solute dynamics
  which makes less assumptions about the solute's short time scale behavior.

  In this work, we are particularly interested in the design of lyotropic liquid 
  crystal (LLC) monomers, a class of amphiphilic molecules whose ordered self-assembled
  phases can be cross-linked into mechanically strong membranes capable of highly 
  selective separations. Inverted hexagonal, H\textsubscript{II}, phase LLC membranes
  are characterized	by hexagonally packed, uniform-sized and straight pores, an ideal
  geometry for high throughput transport. Their pores are lined with the LLC monomer's
  functional groups which can potentially be designed to interact with solutes in a 
  chemically-specific manner. It would be extremely useful to experimentalists if we 
  could connect LLC monomer design with measurable quantities. This goal is complicated
  by the non-Brownian hopping and trapping behavior induced by the membrane's diverse
  topology and inhomogeneous structure result in subdiffusive behaviour out to multiple microseconds of simulation.~\cite{coscia_understanding_2019,coscia_chemically_2019}
  %MRS: to sentence above added to come back to main theme

  We have already advanced our understanding of LLC monomer design in our previous work
  by building stochastic time series models based on the chemical intuition gained
  from qualitative MD studies of solute transport in an H\textsubscript{II} phase LLC
  membrane.~\cite{coscia_chemically_2019,coscia_capturing_2020} We identified three 
  classes of trapping behavior which result in 
  %MRS: tweaked
  subdiffusion out to the microsecond time scale. 
  In our 
  first approach to stochastically model this behavior, we treated the system's dynamics
  as a sequence of anti-correlated hops between power law distributed periods of 
  entrapment, a framework called subordinated fractional Brownian or L\'evy
  motion.~\cite{thiel_weak_2014,teuerle_modeling_2013} In a second approach, we treated 
  solute motion as a Markov state model~\cite{pande_everything_2010} with state-dependent
  dynamics, where we parameterized the state transition probabilities between each of
  eight discrete states, defined by the observed trapping mechanisms, as well as the 
  solute dynamics within each of these states. We showed how one could use realizations
  of any stochastic model like these in order to predict macroscopic flux and selectivity.

  Although powerful and insightful, our previous work required considerable human effort
  in order to analyze the MD trajectories and to hypothesize stochastic models which
  matched the behavior of solutes. It would be significantly more useful and broadly 
  applicable if we could design an approach which 
  %MRS: tweaked to sound more machine-learning-y
  %automatically 
  distinguishes and 
  parameterizes varied solute dynamics 
  in a data-driven way
  and leaves a facile way to generate an ensemble
  of stochastic trajectory realizations that can be used to predict macroscopic transport
  properties.

@@

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but independent observations from 
  the emission distributions associated with each hidden state. Transitions 
  between states are mathematically defined by a matrix of transition probabilities.
  Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm is the most commonly used way to determine the hidden
  state sequence and the underlying parameters of a standard HMM.~\cite{baum_maximization_1970}
  It also possible to achieve the same results using a Bayesian 
  framework.~\cite{scott_bayesian_2002,jasra_markov_2005} 
  %MRS: not entirely clear what purpose of above sentence its - it sort of implied you will use Bayesian framework?
  One can place priors
  on the transition probabilities and parameters of the hidden states' emission
  distributions in order to iteratively infer their values and to estimate the hidden state
  sequence. 
  HMMs have been used in a broad range of applications including speech
  recognition, weather forecasting and bioinformatics.~\cite{juang_hidden_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}.
  In molecular simulations, HMMs have been
  used to characterize molecular kinetics in order to determine the equilibrium 
  distribution of various observables.~\cite{noe_probability_2008}
  %MRS: probably a few more citations that should be added above, since there has been a lot.
  By placing a hierarchical Dirichlet process (HDP) prior on the transition 
  probabilities, it is possible to extend HMMs to an unknown number of hidden
  states, thus allowing the model's complexity to be driven by the complexity
  of the data.~\cite{teh_hierarchical_2006} We will refer to this as the HDP-HMM,
  although in literature it is often called the infinite hidden Markov 
  model.~\cite{beal_infinite_2002} The HDP-HMM elegantly overcomes the main
  shortcoming of standard HMMs, the need to choose the number of states 
  \textit{a priori}, making its use more advantageous in a diversity of 
  applications including economic data and single particle trajectory 
  analysis.~\cite{shi_identifying_2016,hines_analyzing_2015}

  One issue with the original formulation of the HDP-HMM is that it does 
  not distinguish self-transitions and transitions between different states.
  Practically, this can result in rapid switching between states and inferred
  state sequence representations that do a poor job of describing the 
  observations. Fox et al. addressed this issue by adding a `sticky' 
  parameter.~\cite{fox_sticky_2007} The sticky HDP-HMM is useful in situations
  where the user has prior knowledge that state persistence is likely.
  For example, in the case of speaker diarization, it is unlikely that a 
  conversation between two people would rapidly switch between speakers.~\cite{fox_sticky_2011}
  This is also important for our work where we expect state persistence due
  to solute entrapment.

@@

  %MRS: define what ``this'' is - important claim, make it clear what it is you are claiming.
  To our knowledge, this is a novel application and analysis of the sticky 
  HDP-AR-HMM. Typical applications center around determining the number of
  states, the state sequence and occasional analysis of the emission 
  distributions. The addition of temporal correlation to the inference 
  procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015,hamada_modeling_2016}
  Not only do we use the HDP-AR-HMM to segment MD time series into distinct
  dynamical modes, but we cluster and analyze the temporally-correlated 
  parameters of the model in order to learn physical mechanisms and describe
  solute kinetics. Further, we use the model parameters to generate new 
  time series trajectories which enable projections of MD simulations on 
  much longer time scales.

@@

  We ran all MD simulations and energy minimizations using GROMACS 2018.~\cite{bekker_gromacs:_1993,berendsen_gromacs:_1995,van_der_spoel_gromacs:_2005,hess_gromacs_2008}  
  Our Python implementation of the HDP-AR-HMM algorithm is available online at \\
  %MRS: possibly should give the permanent reference to a shirtsgroup fork?  We can discuss which is likely to be more permenant
  \texttt{https://github.com/bencoscia/hdphmm}. All other post-simulation 
  trajectory analysis tools are available online at
  \texttt{https://github.com/shirtsgroup/LLC\_Membranes}.

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10\% water by weight. Approximately one third of the water molecules 
  occupy the tail region with the rest near the pore center.%MRS: cite previous paper here on this.

@@

  %MRS: tweaked
  %We let each solute undergo 
  Each solute system was simulated for 5 $\mu$s of molecular dynamics  simulation. We used a time step of 2 fs
  at a pressure of 1 bar and temperature of 300K controlled by the Parrinello-Rahman 
  %MRS: just wondering, was this semiisotropic, or isotropic?  If isotropic, do we know how different the diagonals of the 
  pressure tensor were (since the system is not totally isotropic)
  barostat~\cite{parrinello_polymorphic_1981} and velocity rescale 
  thermostat~\cite{bussi_canonical_2007} respectively. We recorded configurations every 0.5 ns.

@@

  %MRS: some of this is a repetition of what you said above.
  We implemented the sticky HDP-AR-HMM in Python 
  (\texttt{https://github.com/bencoscia/hdphmm}) which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_nonparametric_2009} Since we do not incorporate
  any novel methodology to the HDP-AR-HMM inference procedure, we refer the 
  interested reader to an excellent and in-depth description of the formulation of
  the HDP-AR-HMM and related models.~\cite{fox_bayesian_2010}

  %\subsubsection*{Parameterizing the Hidden States}\label{method:var_params}
  \textit{Parameterizing the Hidden States}: We describe the dynamics of each 
  state visited by solutes in our MD simulations using a first order vector 
  autoregressive (VAR(1)) model. In general, a VAR($r$) process is characterized by a
  vector of observations in a time series that are linearly dependent on $r$ previous
  values of the time series vector:
  \begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  \label{eqn:var}
  \end{equation}
  Previous observations are weighted by coefficient matrices, $A_i$. The VAR($r$) 
  process is further characterized by a shift in the mean of each dimension by the
  vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
  We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  covariance, $\Sigma$. We limited our analysis to an autoregressive order of $r=1$.
  This means that we only parameterize $A_1$. To simplify notation, we will call it $A$. 
  We used a conjugate matrix-normal inverse-Wishart (MNIW) prior on parameters 
  $A$ and $\Sigma$ and a Gaussian prior on $\mathbf{c}$ in order to infer their 
  values.~\cite{fox_nonparametric_2009}

@@

  We reduced the parameter space using agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage 
  criteria, which works to minimize the sum of the squared differences within all
  clusters.~\cite{ward_hierarchical_1963} We elected to choose the number of clusters
  rather than the distance threshold. For our data, non-parametric methods such as 
  Dirichlet process Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013}
  tend to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information), which could cause the clustered model to combine 
  distinct behavior into a non-physical hybrid state.

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the distance from the 
  pore center and the self-transition probabilities, $T_{ii}$, of each state. $A$ 
  and $\Sigma$ capture the correlation and size of fluctuations. $T_{ii}$ is directly
  related to the average time spent in each state. The distance from the closest pore 
  center is likely meaningful since the membrane is radially inhomogeneous from each 
  pore, transitioning from hydrophilic pores to hydrophobic tails. However, the 
  channels are isotropic along the pore axis so we do not 
  %MRS: added this?  Grammar didn't make sense without.
  cluster
  in means in the $z$ dimension.
  We normalized each clustering dimension by its variance and shifted the mean to zero.

  To take advantage of the system's cylindrical symmetry, we converted the parameters that
  are directly related to solute position into cylindrical coordinates before clustering. 
  We replaced the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ with $\lambda_x$ + 
  $\lambda_y$ and $\lambda_x^2$ + $\lambda_y^2$ respectively, which are invariant with 
  rotation around the $z$ axis. 
  %MRS: maybe reduntant and not quite right, since it reduces the dimensionality of the paraameter su
  % bspace from 3 to 2, but the overall paameter space is bigger than 3
  %in both cases reducing the dimensionality from 3 to 2.  
  We clustered on squared means in $r$, $\mu_r^2 = \mu_x^2 + \mu_y^2$, rather than
  $\mu_x$ and $\mu_y$ individually, with the square $\mu_r$ converting it into the 
  same units as $A$ and $\Sigma$.

@@

  %BJC4: Added
  There are alternative ways to cluster the data dependent on one's goals. Our goal is
  to derive the simplest set of parameter which adequately describe solute motion. 
  However, one may want a more detailed understand of specific dimensions of the 
  clustering dataset. In this case one can cluster on one or more dimensions independently
  and then enumerate the possible clusters. This would ensure very little overlap of
  clusters but may result in a large number of them. Once one has achieved a better 
  understanding of solute-membrane interactions, perhaps by using the HDP-AR-HMM, it may be
  beneficial to incorporate other physically motivated variables into the clustering
  procedure. For example, one can gain a thorough understanding of the dynamics of
  different kinds of trapping behavior related to hydrogen bonding and sodium ion association.
  %MRS: last sentence is a little weird. Audience may be left without knowing how this could be achieved.
  %\subsubsection*{Choosing the Number of Clusters}

@@

  \subsection{Predicting Mean Squared Displacement}\label{method:realizations}
  %BJC4: reworked this section
  %MRS: should explain why here there are two approaches.
  \textit{Generating Stochastic HDP-AR-HMM Trajectory Realizations}: We used two 
  different methods to stochastically generate trajectories:
  \begin{enumerate}[label={Method \theenumi :}, leftmargin=3.5\parindent]
    \item For each independent MD trajectory, we generated stochastic trajectory
    realizations using the finalized parameters of the associated HDP-AR-HMM. This 
    results in 24 separate ensembles of trajectories for each solute studied. 
    \item Using the clustered parameter sets, we generated a single ensemble
	of trajectories. Realizations are representative of the combined behavior
	of all solute trajectories.
  \end{enumerate}

@@

  %MRS: this doesn't make sense - there aren't 1000 trajectories.  You mean method 1?
  For MD trajectories, we bootstrapped according to Method 2.

@@

  Solute flux, as written, can be expressed in terms of the product of solute 
  diffusivity, $D_i$, within the membrane and a sorption coefficient, $K_i$, 
  that quantifies the ability of the solute to move between the bulk solution
  and and the membrane:
  \begin{equation}
    J_i = \frac{D_i K_i}{l}\Delta C
  \end{equation}
  where $l$ is the membrane thickness. For our purposes, $K_i$ and $l$ are 
  irrelevant since we do not study pore entry effects or flux across finite
  length pores. Therefore, in order to calculate selectivity, with respect
  to quantities we can study with our MD simulations, it is only necessary that
  we compare the ratio of diffusion constants:
  \begin{equation}
    S_{ij} = \frac{D_i}{D_j}
    \label{eqn:selectivity_diffusivity}
  \end{equation}
  It can be shown on time scales where motion is uncorrelated, that the diffusion
  constant is linearly proportional to the slope of the MSD curve:~\cite{einstein_investigations_1956}
  \begin{equation}
    \overline{z^2(\tau)} = 2 n D t
    \label{eqn:msd_D}
  \end{equation}
  where $n$ is the number of spatial dimensions (1 in our case). Since the MSD
  %MRS: tweaked
  %of our models 
  of the realizations of our models 
  enters a linear regime, we can estimate diffusion constants 
  %MRS: maybe can say what the timescale of the linear regime, and how much longer the stochastic realizations are?
  based on the slope of the linear region of the predicted MSD curves.

@@

  \textit{Hydrogen bond lifetimes}: We estimated the lifetime of hydrogen bonds by 
  recording the length of sequential frames where solutes remained hydrogen bonded.
  We still counted sequences where hydrogen bonds were broken for a single frame 
  before reforming. Consistent with our previous work, we reported the 95th 
  percentile of hydrogen bond lifetimes since their distribution is not Gaussian
  and to 
  %MRS: 
  %emphasize 
  accentuate
  longer trapping periods.~\cite{coscia_chemically_2019}
